### 2013-03-13 doc/sosp13/memory.tex
edits through the memory section. added consistency model discussion

\begin{figure}[htbp]
  \begin{center}
    \begin{minipage}{\columnwidth}
	\small
	\begin{tabular}{l}
      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
      	\texttt{\scriptsize global\_free( global\_address )} \\ \hline 
      	Allocation in the global heap \\ \\
      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ \hline
      	Memory operations at the home core of a global address \\ \\
      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ \hline
		Perform cache operations to acquire/release global data.  \\
		Acquire copies all data to local node and returns a pointer. \\ 	
		Release optionally copies data back to global memory. \\
	\end{tabular}
      \caption{\label{fig:accessing-memory} Grappa API: accessing memory}     \end{minipage}
  \end{center}
\end{figure}

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:memory-structure} Grappa memory structure}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

@@

\textbf{Explicit caching.} Grappa provides an API to fetch a global
pointer of any length and return a local pointer to a cached copy of the
global memory.  Grappa cache operations have the usual read-only and
read-write variants, along with a write-only variant used to initialize
data structures. Languages for distributed shared memory systems have
done optimizations to achieve a similar goal. For example, the UPC
compiler coalesces struct and array accesses into remote get/put
\cite{Chen:2005}, and Fortran D compiler's message vectorization hoists
small messages out of a loop \cite{FortranD:1992}. Caching in Grappa
additionally provides a mechanism for exploiting temporal locality by
operating on the data locally.

Under the hood, Grappa performs the mechanics of gathering chunks of
data from multiple system nodes and presenting a conventional appearing
linear block of memory as a local pointer into a cache. The strategy
employed is to issue all the constituent requests of a cache access
request  (as Active Messages) and then yield until all responses have
occurred.  Currently, Grappa caches are \emph{not} coherent, requiring
the programmer to maintain consistent access to data.  Future work will
develop a software directory based coherence scheme to simplify
consistent access to global data.

\textbf{Delegate operations.} When the access pattern has low-locality,
it is more efficient to modify the data on its home core rather than
bringing a copy to the requesting core and returning it after
modification. Delegate operations provide this capability. Applications
can dispatch computation to be performed on individual machine-word
sized chunks of global memory to the memory system itself (e.g.,
\emph{fetch-and-add}).  Delegate operations, proposed in
\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the
primary synchronization method in Grappa.

Delegate operations are always executed at the home core of their
address, and while arbitrary memory operations can be delegated, we
restrict the use of delegate operations in three ways to make them more
useful for synchronization. First, we limit each task to one outstanding
delegate operation to avoid the possibility of reordering in the
network. Second, we limit delegate operations to operate on objects in
the 2D address space or objects that fit in a single block of the linear
address space so they can be satisfied with a single network request.
Finally, no context switches are allowed while the data is being
modified. Given these restrictions, we can ensure that delegate
operations for the same address from multiple requesters are always
serialized through a single core in the system, providing atomic
semantics without using atomic operations. 
Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
cache operations interact.

\begin{figure*}[htb]
\begin{center}
  \includegraphics[width=1.5\columnwidth]{figs/delegate-cache}
\begin{minipage}{1.9\columnwidth}
  \caption{\label{fig:delegate-cache} \textbf{Delegation and cache example:} In step 1, a core allocates an array in
the global heap. It then spawns two tasks on remote cores to increment
elements of the array.  The first task increments two elements of the array using cache
operations. In step 2, the task is invoked. A cache request is issued
for two adjacent integers starting at the second element of the
array. Since these element are stored in the memories of two different
cores, this requires the sending of two messages in step 3. The task
is suspended until both responses arrive in step 4. The data carried
in these responses is stored in the local buffer. The elements are
then increment in the buffer. Then in step 5, the modified data is
sent back to the home node. Acknoledgements are returned in step 6 so
the task knows when the writes are complete. The second task increments an element of the array with a delegate
operation. In step 7, the task is invoked. A delegate request is sent
to the home core of the array element with the increment value. The
task suspends until the response is received. In step 8, the increment
is executed on the remote core. A response is returned in step 9 with
the previous value of the array element.
}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure*}

>>>>

\begin{figure}[htbp]
  \begin{center}
    \begin{minipage}{\columnwidth}
	\small
	\begin{tabular}{l}
      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
      	\texttt{\scriptsize global\_free( global\_address )} \\ 
      	Allocation in the global heap \\ \\
      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ 
      	Memory operations at the home core of a global address \\ \\
      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ 
		Perform cache operations to acquire/release global data.  \\
		Acquire copies all data to local node and returns a pointer. \\ 	
		Release optionally copies data back to global memory. \\
	\end{tabular}
      \caption{\label{fig:accessing-memory} Grappa API for memory accesses}     \end{minipage}
  \end{center}
\end{figure}

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:memory-structure} Global memory referencing in Grappa}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

@@

\textit{Explicit caching.} Grappa provides an API to fetch a global pointer of
any length and return a local pointer to a cached copy of the global memory.
Grappa cache operations have the usual read-only and read-write variants,
along with a write-only variant used to initialize data structures. Languages
for distributed shared memory systems have done optimizations to achieve a
similar goal. For example, the UPC compiler coalesces struct and array
accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's
message vectorization hoists small messages out of a loop
\cite{FortranD:1992}. Caching in Grappa additionally provides a mechanism for
exploiting temporal locality when available by operating on the data locally.

Under the hood, Grappa performs the mechanics of gathering chunks of data from
multiple system nodes and presenting a conventional appearing linear block of
memory as a local pointer into a cache. The strategy employed is to issue all
the constituent requests of the desired block (as Active Messages) and then
yield until all responses have occurred. Currently, Grappa caches are
\emph{not} kept coherent automatically. If that is desired, the programmer to
is responsible for pushing the updated block back to its home node. Given that
synchronization is explicit in Grappa, we don't expect this to be a
significant burden (programmers can push updates before a relevant
synchronization operation).

% Future work will develop a software directory based coherence scheme to
% simplify consistent access to global data.

\textit{Delegate operations.} When the access pattern has low-locality,
it is more efficient to modify the data on its home core rather than
bringing a copy to the requesting core and returning it after
modification. Delegate operations provide this capability. Applications
can dispatch computation to be performed on individual machine-word
sized chunks of global memory to the memory system itself (e.g.,
\emph{fetch-and-add}).  Delegate operations, proposed in
\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are also the
primary synchronization method in Grappa.

Delegate operations are always executed at the home core of their
address, and while arbitrary memory operations can be delegated, we
restrict the use of delegate operations in three ways to make them more
useful for synchronization. First, we limit each task to one outstanding
delegate operation to avoid the possibility of reordering in the
network. Second, we limit delegate operations to operate on objects in
the 2D address space or objects that fit in a single block of the linear
address space so they can be satisfied with a single network request.
Finally, no context switches are allowed while the data is being
modified. Given these restrictions, we can ensure that delegate
operations for the same address from multiple requesters are always
serialized through a single core in the system, providing atomic
semantics without using actual atomic operations (and thus avoiding their typical high cost). 
Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
cache operations interact.

\begin{figure*}[htb] \begin{center}
\includegraphics[width=1.5\columnwidth]{figs/delegate-cache}
\begin{minipage}{1.9\columnwidth} \caption{\label{fig:delegate-cache}
\textbf{Delegation and cache example:} A core allocates an array in the global
heap (1). It then spawns two tasks on remote cores to increment elements of
the array. The first task increments two elements of the array using cache
operations. The first task is then invoked (2). A cache request is issued for two
adjacent integers starting at the second element of the array. Since these
element are stored in the memories of two different cores, this requires two messages to be sent (3). The task is suspended until both responses arrive
(4). The data carried in these responses is stored in the local buffer. The
elements are then incremented in the buffer. The modified data is then sent back
to the home node (5). Acknowledgements are returned (6) so the task
knows when the writes are complete. The second task increments an element of
the array with a delegate operation. The task is invoked (7). A delegate
request is sent to the home core of the array element with the increment
value. The task suspends until the response is received. The increment
is executed on the remote core (8). A response is returned (9) with the
previous value of the array element. } \end{minipage} \vspace{-3ex}
\end{center} \end{figure*}

\paragraph{Memory consistency model discussion} As mentioned earlier, all
synchronization operations are done via delegate operations. Since they all
execute in their home node, they are guaranteed to be serialized , with their
updates visible to all cores across the system in the same order. Also, tasks
are restricted to a single outstanding delegate operation at a time, and
therefore are not subject to potential request reordering. Consequently, all
synchronization operations execute in program order and are made visible in
the same order to all cores in the system. These properties are sufficient to
guarantee a memory model that offers sequential consistency for data-race-free
programs~\cite{drf} (all accesses to shared data are separated by
synchronization). This is the memory model that underpins C/C++.

Note, however, that if the application code uses explicit caching on shared
data, all updates must be sent back to the home node before the
synchronization operation that protects the data is performed.
